---
title: "RI for LATE (under construction)"
output: github_document
vignette: >
  %\VignetteIndexEntry{Anderson-Rubin Permutation Tests for Noncompliance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

**Author:** Arya Gadage and Haoge Chang<br/>
**Last updated:** `r format(Sys.Date(), "%B %d, %Y")`

# Introduction

This package implements the randomization-based inferential procedure for the Local Average Treatment Effect (LATE) developed in [[2]](https://academic.oup.com/biomet/article/113/2/asag010/8487895) and [[4]](https://academic.oup.com/jrsssa/article/168/1/109/7084141).


## Overview

In randomized experiments, participants do not always take the treatment they were assigned to — the treatment they actually receive can differ from their assignment. This gap is called **noncompliance**. When it is present, a natural target of inference is the **Local Average Treatment Effect (LATE)** [[3]](https://www.jstor.org/stable/2951620): the average causal effect of treatment among **compliers**, the subpopulation whose treatment status changes as a result of treatment assignment.

To state this precisely, consider a sample of individuals $i = 1, \ldots, n$, and for each let

- $Y_i(1), Y_i(0)$ — the potential outcomes under treatment and control;
- $D_i(1), D_i(0)$ — the treatment actually taken when assigned to treatment and to control.


A complier is an individual with $D_i(1) > D_i(0)$: they take the treatment when assigned to it, but not otherwise. We use $\mathcal{C}=\{i \,:\, D_i(1) > D_i(0)\}$ to denote the set of compliers. The LATE is the average treatment effect over this group,

$$
\tau_{\text{LATE}} \;=\; \frac{1}{n_c} \sum_{i \in \mathcal{C}} \bigl( Y_i(1) - Y_i(0) \bigr),
$$

where $n_c = |\mathcal{C}|$ is the number of compliers. This package provides a randomization-based inferential procedure for $\tau_{\text{LATE}}$ with two complementary guarantees:

- **Tests and p-values** are *finite-sample valid* for testing the sharp null $Y_i(1) = Y_i(0)$ for all $i \in \mathcal{C}$, and *asymptotically valid* for testing $\tau_{\text{LATE}} =0$ (under treatment-effect heterogeneity).
- **Confidence sets** are *finite-sample exact* under treatment-effect additive homogeneity ($Y_i(1) = Y_i(0) + \tau$ for all $i \in \mathcal{C}$, with $\tau$ not depending on $i$), and *asymptotically valid* for the LATE under heterogeneity.
The procedures are robust against low compliance rates.

The package supports covariate adjustment via linear regression, which can improve precision and guard against chance covariate imbalance.

# (Important) Complete Randomization Design

The package currently supports completely randomized designs with a binary treatment. The remaining statistical regularity conditions for asymptotic g
uarantees are stated in [[2]](https://academic.oup.com/biomet/article/113/2/asag010/8487895).

# Note
 1. The package includes two algorithms that implement the same inferential procedure but differ in speed. Algorithm 2 is faster and is used by default; Algorithm 1 is retained for comparison and validation.
 2. The authors used Claude to migrate the original code from [[2]](https://academic.oup.com/biomet/article/113/2/asag010/8487895) into this repository and to help design the test cases.


## Demonstration

# Install packages

`rilate` is not yet on CRAN. Install the development version from
[GitHub](https://github.com/hchang1508/rilate) with:

``` r
# install.packages("remote")
remotes::install_github("hchang1508/rilate")
```

Then load it with `library(rilate)`, as in the example below.

# Synthetic Data Example

The workflow has three steps: simulate a randomized experiment with
noncompliance, (optionally) attach baseline covariates, and call `run_AR()`.

**1. Simulate the experiment.** `gen_sim_data()` draws a completely randomized
experiment from a population of always-takers, compliers, and never-takers. Here
the true complier effect (the LATE) is `1`.

```{r sim-data}
library(rilate)

set.seed(7)
sim <- gen_sim_data(
  N         = 800,
  N1        = 400,                  # 400 assigned to treatment, 400 to control
  fractions = c(0.20, 0.50, 0.30),  # always-takers, compliers, never-takers
  taus      = c(0, 1, 0),           # principal-stratum effects; complier LATE = 1
  mode      = "constant"
)
dat <- as.data.frame(sim$observed)  # columns: Y_observed, D_observed, assignment
```

**2. Attach prognostic covariates.** `gen_sim_data()` returns only the outcome,
treatment, and assignment, so we append two baseline covariates by hand. They are
drawn independently of assignment (randomization keeps them balanced across arms)
but drive part of the outcome, so adjusting for them removes noise and sharpens
inference.

```{r add-covariates}
n <- nrow(dat)
dat$x1 <- rnorm(n)
dat$x2 <- rnorm(n)
dat$Y_observed <- dat$Y_observed + 2 * dat$x1 - 1.5 * dat$x2

head(dat)
```

**3. Run the randomization inference procedure.** `y`, `d`, and `z` name the
outcome, treatment-received, and assignment columns (shown here at their
defaults). Passing `x` runs *both* an unadjusted analysis and a
covariate-adjusted one. Each returns a confidence set for the LATE and a
randomization p-value for the sharp null ($Y_i(1)=Y_i(0)$ for all compliers) and
weak null ($\tau_{\text{LATE}} = 0$). (`run_AR()` also prints a human-readable
report to the console; it is suppressed here.)

```{r run-ar}
invisible(capture.output(
  res <- run_AR(dat,
                y = "Y_observed", d = "D_observed", z = "assignment",
                x = c("x1", "x2"), algorithm = "algo2",
                n_rand = 1000, seed = 1)
))
```

The result carries a `without`- and a `with`-covariates entry, each with a
`confidence_set` and a `p_value`:

```{r summarise}
fmt_cs   <- function(cs) paste(vapply(cs, function(iv)
  sprintf("[%.3f, %.3f]", iv[1], iv[2]), character(1)), collapse = " U ")
cs_width <- function(cs) sum(vapply(cs, function(iv) diff(iv), numeric(1)))

data.frame(
  analysis = c("without covariates", "with covariates"),
  p_value  = round(c(res$results$without$p_value, res$results$with$p_value), 4),
  CS_95    = c(fmt_cs(res$results$without$confidence_set),
               fmt_cs(res$results$with$confidence_set)),
  width    = round(c(cs_width(res$results$without$confidence_set),
                     cs_width(res$results$with$confidence_set)), 3),
  row.names = NULL
)
```

Both 95% confidence sets cover the true LATE of `1`. Adjusting for the two
prognostic covariates shrinks the confidence set roughly threefold (from about
`[-0.47, 1.49]` to `[0.83, 1.54]`) and sharpens the p-value for
$\tau_{\text{LATE}} = 0$ from about `0.27` — no detectable effect — to about
`0.001`. This is the precision gain covariate adjustment buys when the covariates
are prognostic for the outcome.

# Assess Coverage
# Compare Two algorithms (time )

# References

[1] Anderson, T. W., & Rubin, H. (1949). Estimation of the parameters of a single equation in a complete system of stochastic equations. *Annals of Mathematical Statistics*, 20(1), 46-63.

[2] Aronow, P. M., Chang, H., & Lopatto, P. (2026). Randomization-Based Confidence Sets for the Local Average Treatment Effect. *Biometrika*, 113(2).

[3] Imbens, G. W., & Angrist, J. D. (1994). Identification and Estimation of Local Average Treatment Effects. Econometrica, 62(2), 467-475.

[4] Imbens, G. W., & Rosenbaum, P. R. (2005). Robust, accurate confidence intervals with a weak instrument: quarter of birth and education. *Journal of the Royal Statistical Society: Series A*, 168(1), 109-126.


